---
title: "cyclist_analysis_2024_01_09"
author: "Alisson Vinicius Salvador de Lima"
date: "2024-01-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis

In this report, it will be done some resumes, for further analysis in dashboards. All resumes will ever try to frame different possible aspects that define to different rider profiles - member riders or casual riders.

# Preparing the datasets

## Install Packages

```{r install}

install.packages("skimr") # Cleaning
install.packages("tidyverse") # Data manipulation
install.packages("janitor") # Cleaning
install.packages("stringr") # Cleaning strings
install.packages("lubridate") # Date manipulation
```

## Loading Packages

```{r load packages}

library(tidyverse)
library(skimr)
library(janitor)
library(dplyr)
library(tidyr)        
library(ggplot2)
library(stringr)
library(lubridate)

```

## Load Dataset

Here we load the dataset previously processed on previous report "cyclist_cleaning_2024_01_08.Rmd".

```{r load data}

# for general analysis
df_analysis_01 <- read.csv("data/processed_data_2024_01_08.csv")

# for hour analysis
df_hour <- read.csv("data/df_for_hour_analysis.csv")

```

## Check Integrity

```{r glimpse}

glimpse(df_analysis_01)

```

# Summarizing

## Summarize Ride Length and Distance Values

Count members and casual users

```{r group and summarize}

member_casual_count <- df_analysis_01 %>%
  group_by(member_casual) %>%
  summarise(count = n())

```

## Descriptive analysis for all data as a whole

```{r group and summarize}

summary_all <- df_analysis_01 %>%
  summarise(average_lenth=mean(ride_length), 
            min_length=min(ride_length),
            max_length=max(ride_length),
            st_length= sd(ride_length),
            average_distance=mean(distance),
            min_distance=min(distance),
            max_distance=max(distance),
            st_distance= sd(distance))

```

## Descriptive analysis for each type rider profile (member or casual) 

```{r group and summarize}

summary_member_casual <- df_analysis_01 %>%
  group_by(member_casual) %>%
  summarise(average_lenth=mean(ride_length),
            min_length=min(ride_length),
            max_length=max(ride_length),
            st_length= sd(ride_length),
            average_distance=mean(distance),
            min_distance=min(distance),
            max_distance=max(distance),
            st_distance= sd(distance))

```

## Summary Table

Just formatting a new table to fit the antecedent table

```{r descriptive merge}

d <- data.frame(member_casual = c("both"))

summary_all <- cbind(d, summary_all)

summary <- rbind(summary_all, summary_member_casual)

```

```{r summary table}

View(summary)

```


## Let us check out the length distribution over member and casual riders on a boxplot

Check length

```{r boxplot len} 

ggplot(df_analysis_01, aes(x = member_casual, y = ride_length)) +
  geom_boxplot()

```
Check distance

```{r boxplot dist} 

ggplot(df_analysis_01, aes(x = member_casual, y = distance)) +
  geom_boxplot()

```
## Distribuition

Check Length

```{r histogram len} 

ggplot(df_analysis_01, aes(x = ride_length)) +
  geom_histogram() +
  facet_wrap(~member_casual, ncol = 2)


```
Check distance

```{r histogram dist} 

ggplot(df_analysis_01, aes(x = distance)) +
  geom_histogram() +
  facet_wrap(~ member_casual, ncol = 2)

```


## Summarize Information by Day

```{r sumarize day}

summary_day <- df_analysis_01 %>%
  group_by(member_casual, day_of_week, ride_month, ride_day) %>%
  summarise(average_lenth=mean(ride_length),
            min_length=min(ride_length),
            max_length=max(ride_length),
            average_distance=mean(distance),
            min_distance=min(distance),
            max_distance=max(distance),
            daily_rides = n())

```
## Summarize the number of rides by day and rideable type

```{r sumarize rideable type}

summary_rideable_type <- df_analysis_01 %>%
  group_by(member_casual,rideable_type, day_of_week, ride_month, ride_day) %>%
  summarise(average_lenth=mean(ride_length),
            average_distance=mean(distance),
            type_rides_count = n())

```
## Summarize by Weekday and Hour
```{r}

summary_hour <- df_hour %>%
  group_by(member_casual, day_of_week, ride_hour) %>%
  summarise(average_length = mean(ride_length),
    hour_ride_count = n())

```


## Get the month names

These transformations are optional

```{r transform month}

summary_day$ride_month <- month.name[summary_day$ride_month]

summary_rideable_type$ride_month <- month.name[summary_rideable_type$ride_month]

```

## Export the tables to be analysed in dinamic tables

```{r export}

write.csv(summary_day, "data/summary_data_count_2024_01_12.csv", row.names = FALSE)
write.csv(summary, "data/summary_data_length_2024_01_10.csv", row.names = FALSE)
write.csv(summary_rideable_type, "data/summary_data_rideable_type.csv", row.names = FALSE)
write.csv(summary_hour, "data/summary_hour.csv", row.names = FALSE)

```
